---
title: "Getting Started with sublime"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with sublime}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## Introduction

'sublime` implements some sub-data selection and randomized
sketching methods for speeding up linear regression, feature selection and 
generalized linear models on large datasets. Rather than fitting a 
model on all $N$ observations, each method constructs a
smaller subset of size $k \ll N$ and fits the model on that subset 
instead, reducing the computational cost.

The package provides five functions:

| Function     | Model                       | Idea                                                            |
|--------------|-----------------------------|-----------------------------------------------------------------|
| `IBOSS()`    | Linear regression           | Deterministically pick rows with extreme covariate values       |
| `GenIBOSS()` | GLMs (Logistic, Poisson, …) | IBOSS applied to an information-weighted design                 |
| `SRHT()`     | Linear regression           | Randomized Hadamard sketch, then subsample                      |
| `CLASS()`    | Lasso                       | Repeated subsampled Lasso + IBOSS to select variables and rows  |
| `HTCLASS()`  | Lasso                       | `CLASS()` on a Hadamard-transformed matrix                      |

This vignette walks through each function with a small example.

```{r}
library(sublime)
```

## IBOSS: Information-Based Optimal Subdata Selection

`IBOSS()` is an algorithm for ordinary linear regression on very large $N$.
It selects the $k$ rows of $X$ that maximize a lower bound on the
determinant of the Fisher information matrix, by repeatedly picking
observations with extreme values in each predictor.

```{r}
set.seed(42)
n <- 20000
p <- 5

X <- matrix(rnorm(n * p), ncol = p)
beta <- c(2, -1, 0.5, 3, -2)
y <- as.numeric(X %*% beta + rnorm(n))

res <- IBOSS(X = X, y = y, k = 500)

fit <- lm(res$y_selected ~ res$X_selected)
coef(fit)
```

`IBOSS()` can also be pointed at a CSV file directly, with the last
column treated as the response:

```r
res <- IBOSS(csv = "mydata.csv", k = 500, header = TRUE)
```

If `intercept = TRUE`, the first column of `X` is treated as an intercept
term and excluded from the row-selection criterion (but still included
in the returned subdata).

## GenIBOSS: IBOSS for Nonlinear Models

`GenIBOSS()` first fits the model on a small pilot subsample of
size `nSample`, using the result to build an
information-weighted design, and then running IBOSS-like selection on
that weighted design.

```{r}
set.seed(42)
n <- 20000
p <- 4

X <- matrix(rnorm(n * p), ncol = p)
beta <- c(1, -1, 0.5, 2)
eta <- as.numeric(X %*% beta)
prob <- 1 / (1 + exp(-eta))
y <- rbinom(n, 1, prob)

fit <- GenIBOSS(
  X = X,
  y = y,
  nSample = 500,
  k = 1000,
  family = binomial(),
  add_logs = TRUE
)

coef(fit$final_model)
```

With `add_logs = TRUE`, `GenIBOSS()` also fits a final `fastglm` model on
the selected subdata and returns it as `final_model`; with
`add_logs = FALSE` (the default) it returns only the selected
`X_selected` / `y_selected`, leaving the final fit to the user.

Any standard `family` object works, as long as it defines `linkinv`,
`mu.eta`, and `variance` — e.g. `gaussian()`, `binomial()`, `poisson()`.

## SRHT: Subsampled Randomized Hadamard Transform

`SRHT()` is a randomized sketching alternative to IBOSS. It first applies a random
Hadamard-based rotation to the whole design (which tends to equalize
leverage scores across rows), and then draws a uniform subsample of size
`k` from the transformed data. This reduces the computational cost for 
linear regression on the result.

```{r}
set.seed(42)
n <- 20000
p <- 5

X <- matrix(rnorm(n * p), ncol = p)
beta <- c(2, -1, 0.5, 3, -2)
y <- as.numeric(X %*% beta + rnorm(n))

res <- SRHT(X = X, y = y, k = 2000, intercept = TRUE)
str(res)
```

If `intercept = TRUE` and the first column of `X` is not already a
column of ones, `SRHT()` prepends one before sketching.

## CLASS: Combined Lasso And Subdata Selection

`CLASS()` targets datasets with a large number of features, where variable
selection (via Lasso) is itself expensive on the full data. It repeatedly
draws small uniform subsamples of size `nSample`, fits Lasso on each via
`glmnet::cv.glmnet()`, and counts how often each variable is selected
across `nTimes` repetitions. K-Means clustering is done on the resultant
frequencies, with k=2. The variables corresponding to the higher mean are chosen,
the rest are discarded. We then perform IBOSS on the resultant dataset, after
which linear regression is performed on the sub-data.

```{r}
set.seed(42)
n <- 50000
p <- 50

X <- matrix(rnorm(n * p), ncol = p)
beta <- c(rep(2, 5), rep(0, p - 5))
y <- as.numeric(X %*% beta + rnorm(n))

res <- CLASS(
  X = X,
  y = y,
  nSample = 1000,
  nTimes = 50, 
  k = 2000
)

#The originial algorithm uses an nTimes of 100. We use a smaller nTimes
#here for illustrative purposes.

res$selected_indices  # active variables
res$beta_final         # full-length coefficient vector (zeros for excluded vars)
res$mse
res$r_squared
```
`CLASS()` returns a list containing the selected subdata
(`X_f`, `y_f`), the fitted intercept and coefficients, the indices of the
variables selected as active, the raw selection frequency counts, and
in-sample MSE / $R^2$ computed on the full dataset using the final
coefficients.

## HTCLASS: CLASS + Hadamard Transform

`HTCLASS()` follows the same procedure as `CLASS()`, but first applies a
Hadamard transform (`HT()`) to `X` and `y` before running the subsampled
Lasso stage. Preconditioning with a Hadamard matrix spreads out
leverage/influence across rows, which can improve accuracy relative to
plain `CLASS()` on datasets where a few rows dominate the OLS fit.

```{r}
res <- HTCLASS(
  X = X,
  y = y,
  nSample = 1000,
  nTimes = 50,
  k = 2000
)

res$beta_final
res$mse
```

The return value has the same structure as `CLASS()`.

## References

IBOSS: H. Wang, M. Yang, and J. Stufken, "Information-based optimal
subdata selection for big data linear regression," *Journal of the
American Statistical Association*, vol. 114, no. 525, pp. 393-405, 2019.
<https://doi.org/10.1080/01621459.2017.1408468>

GenIBOSS: J. Yu, J. Liu, and H. Wang, "Information-based optimal subdata
selection for non-linear models," *Statistical Papers*, vol. 64, no. 4,
pp. 1069-1093, Aug. 2023.
<https://doi.org/10.1007/s00362-023-01430-3>

CLASS: R. Singh and J. Stufken, "Subdata selection with a large number of
variables," *The New England Journal of Statistics in Data Science*, vol.
1, no. 3, pp. 426-438, 2023. <https://doi.org/10.51387/23-NEJSDS36>

SRHT: J. A. Tropp, "Improved analysis of the subsampled randomized
Hadamard transform," *Adv. Adapt. Data Anal.*, vol. 3, no. 1-2, pp.
115-126, 2011. <https://doi.org/10.1142/S1793536911000787>
